NDNS multilevel LCA CW2CB2 or CW2CB3

### 2018-08-28 paper/Chapters/Chapter2.tex
language checked by Hunter

In the diary, participants were asked to record portion sizes in household measures (e.g. one tablespoon of beans, one Kit Kat finger-size), or for packaged foods to note the weight indicated on the packet. For homemade dishes, participants were asked to record on a separate page in the diary the individual ingredients and quantities for the whole dish along with a brief description of the cooking method and how much of dish they had consumed. In each eating occasion, in addition to the details of what and how much was eaten, participants were also asked to record: when was it, where they were, and who they were eating with. An example, used as guidance for participants, of a food diary for one day is shown in \textbf{Appendix \ref{AppendixE}}.\vspace{-0.3cm}

@@

\begin{itemize}
	\item Total energy intake = (protein(gram) $\times$ 17) + (fat(gram) $\times$ 37) + (carbohydrate(gram) $\times$ 16) + (alcohol(gram) $\times$ 29)  kJ;
	\item Carbohydrate intake = total sugars (gram) + starch (gram); 
%	\item Total sugars = sum of all individual sugars (gramme).
\end{itemize}

Time across a typical survey day was divided into 7 time slots in the dietary diary of NDNS RP: 6 am to 9 am, 9 am to 12 noon, 12 noon to 2 pm, 2 pm to 5 pm, 5 pm to 8 pm, 8 pm to 10 pm, and 10 pm to 6 am next morning. To produce a sequence of discrete responses regarding the carbohydrate intake we are interested, the energy consumption within each time slot over the four days of survey for each participant were calculated. The percentages of total energy intake that contributed by carbohydrate within each time slot were then estimated. Since we planned to apply latent class analysis (LCA) in the current study, in which the observed indicators for latent classes must be categorical, the responses were then dichotomised according to the carbohydrate contribution to the energy intake at cut-off value of 50\%, i.e. if within a time slot where there is any energy intake occurred, carbohydrate consumption was categorised into whether it's energy contribution was lower or higher/equal to 50\% of total energy intake within that time slot. Consequently, for each day of the recording, there were 7 data points generated by the diary, each data point included one of the following responses:

@@

\begin{itemize}
	\item Randomly select PSUs from the Postcode Address File; 
	\item Randomly select 27 or 28 addresses in that postcode area; 
	\item Randomly select one household at that address; 
	\item Selected addresses were randomly allocated to one of two groups to determine whether an adult (aged 19 years or older) and a child (aged 1.5 to 18 years) or a child only, were selected for interviews.
\end{itemize}
\vspace{-0.6cm}
%-----------------------------------
%	SUBSECTION 2
%-----------------------------------
\subsection{Response rates}\vspace{-0.3cm}

@@

The strata that used to calibrate proportions in the sample include: age-group (1.5-3, 4-6, 7-10, 11-15, 16-18, 19-24, 25-29, 30-39, 40-49, 50-59, 60-64, 65-69, and over 70 years); sex (men or women); and regions (Northern Ireland, Scotland, Wales, and the nine regions of England).

Two steps of weighting system are designed in the NDNS RP to assure that the combined sample will be representative of the UK population:

\begin{enumerate}
	\item An overall selection weight, which is the product of the address, dwelling unit, catering (household) unit, and individual selection weights, was generated to correct for the unequal selection probabilities. These weights are the inverses of the selection probabilities at each level of the random sampling process and they can be used to compensate for differences in the chance of selection of an individual.
	\item An iterative procedure was used to adjust the selection weights until the distribution of the weighted sample matched that of the population for age-group, sex, and geographical region. Population distributions were taken from the mid-year population estimates \parencite{OfficeofNationalStat}. 
\end{enumerate}

Another two sets of weights were generated to correct for differential non-response (either due to refusal or inability) to 1) nurse visit and 2) giving a blood sample. Response rates to the nurse visit among those completed a dietary diary was approximately 75\%, to the blood sample in adults were 51\%, 57\%, and 50\% for years 1 to 4, 5 to 6, and 7 to 8, respectively. In creating the nurse/blood sample weight, a logistic regression model was used by the NDNS RP study team to model the relationship between response to nurse visit/giving a blood sample and a set of predictor variables (socio-demographic, participant and catering/household unit characteristics). The model generated a predicted probability for each participant, which is the probability would agree to a nurse visit/provide a blood sample, given the characteristics of the individual and the household unit. Participants with characteristics associated with non-response were under-represented in the sample and therefore receive a low predicted probability. The inverses of these predicted probabilities were used as a set of non-response weights so that participants with a low predicted probability got a larger weight, increasing their representation in the sample. Then the nurse/blood sample weights were re-scaled so that the sum of the weights equalled the number of participants who had a nurse visit/who provided a blood sample. The final nurse/blood weights should, therefore, make the sample participants representative of all eligible persons in the population.

@@

Computer-assisted personal interviews were conducted for the selected individuals by trained interviewers to collect background information on smoking habits (current, ex-smokers, and never), ethnicity (white, non-white), education level (lower than degree/degree or above level), living with a partner or not and other socio-demographic variables. Participants also had their height, weight, blood pressure, waist circumferences (WC) measured by the nurses.

Specifically, blood pressure was measured in a sitting position using an automated, validated machine, the Omron HEM907, after a five-minute rest. The means of second and third readings, taken at one-minute intervals, were used in the current report. Hypertension was defined as with systolic blood pressure of 140 mmHg or above, and/or diastolic blood pressure of 90 mmHg or above, and/or taking any medication specifically to reduce blood pressure.

@@

Latent class analysis is a statistical technique that identifies categorical latent (unobserved) class variables on the basis of observed categorical variables \parencite{collins2010latent}. It belongs to the family of latent variable models and is directly analogous to the factor analysis model. The major difference is that the latent variable in LCA is categorical, not continuous as in factor analysis. The basic assumptions in LCA are independent observations, and local independence, as shown in the fundamental expression of a typical LCA model: \vspace{-0.8cm}

@@

\textbf{Equation \ref{LCA}} indicates that responses for an observation to the measuring variables are independent of one another given its membership in latent class $t$. However, in the NDNS RP data set, the assumption of independent observations is violated. Each individual completed their dietary diary for three/four consecutive days, their diary recordings were later converted into three/four sequences of categorical responses reflecting the type of carbohydrate consumption at each time slot of the day. The observed sequences (observed days) are nested within the participants and therefore are not independent. This nested data structure requires multilevel techniques. 
\vspace{-0.3cm}

@@

Multilevel latent class analysis accounts for the nested structure of the data by allowing latent class intercepts to vary across level 2 units and thereby examining if and how level 2 units influence the level 1 latent classes. These random intercepts allow the probability of membership in a particular level 1 (observation days) latent class to vary across level 2 units (e.g., here in the current context are the individuals). Essentially this allows the probability that an observation day will belong to a
particular day-level latent class to vary across individual-levels.\vspace{-0.5cm}

@@

Thus, for a binary outcome $C_{ij} = 0, 1$ (low $=0$ or high $=1$ carbohydrate eating days), where $i$ denotes the observation days $(i = 1, 2, 3, 4)$, and $j$ denotes the individual $(j = 1, 2, \cdots, 6155)$. The 2-level random intercept logistic regression model can be expressed as:\vspace{-0.4cm}

@@

The same framework can be used to consider random effects in an LCA model, but instead of saying that $C_{ij}$ is either low or high carbohydrate eating days as if we already know, it is now replaced by a latent variable $G_{ij}$ which indicates the typologies of carbohydrate eating patterns. Then we can use the day-level data to assess the log-odds of belonging to the $k$th type of carbohydrate eating pattern on a specific day of the survey, and we allow the log-odds to vary across individuals. Therefore, for some persons, the log-odds of having a $k$th type of carbohydrate eating pattern during the survey can be high, but for the other persons, the log-odds of following the $k$th type of carbohydrate eating pattern can be low.

@@

\begin{equation}
P(C_{ij} = t | CB_j = m)  = \frac{\exp(\gamma_{tm})}{\sum_{r=1}^{T}\exp(\gamma_{tm})}
\end{equation}

@@

\begin{itemize}
	\item Firstly, we ignored the multilevel structure of the data and estimated a series of traditional LC models to determine the number of classes at the observational day-level;
	\item Next, a series of MLCA models were fitted to account for the multilevel structure of the data. In these models, the number of day-level classes was based on the best fitting LCA model from the first step, and the LCA model at the individual-level was estimated to identify the number of individual-level latent classes;
	\item Thirdly, when the number of individual-level latent classes is defined based on the previous stage, the number of day-level classes was modified (one class lower and one class higher than in the second step) to see the effect of changing level 1 classes and confirm the best fitting model.
\end{itemize}

The number of classes in level 1 was determined by 1) the evaluation of model fit indices, including the Bayesian information criterion (BIC), and entropy which is a statistic that summarizes latent class probabilities where values near 1 indicate better latent class separation; 2) the Lo-Mendell-Rubin Likelihood Ratio Test (LMR-LRT) \parencite{lo2001testing, nylund2007deciding} which compares $q$ vs. $q-1$ classes models, where $q$ is the number of latent classes and most importantly, 3) pattern interpretability. In the steps of performing multilevel LCA, where LMR-LRT is not available, same rules of model fit indices and pattern interpretability were used to determine the optimal combination of latent classes in observation day-level and individual-level. MLCA models were fitted in Mplus 7.4 \parencite{muthen2005mplus}, the Mplus syntax and outputs are shown in \textbf{Appendix \ref{AppendixB}}. \vspace{-0.3cm}

@@

day-level latent classes identified by the first step of MLCA were tabulated according to the day of week and also whether the diary was recorded during weekends or not. A contingency table giving the frequency of responses across the 7-time slots of the survey days was produced. Descriptive statistics for the dietary day-level recordings according to the latent class memberships were presented. Pearson $\chi^2$ test was used to compare the distribution of categorical variables. One-way Analysis of Variance (ANOVA) was used to compare the means across the multiple groups for continuous variables.

@@

\begin{enumerate}
	\item The crude association between the carbohydrate eating groups and the outcomes was first examined. 
	\item Potential confounders of the association between carbohydrate eating groups (exposure) and the outcomes were selected depending on the descriptive statistical analyses conducted above, i.e. those are associated with both the exposure and the outcome and also not on the causal pathway were selected as potential confounders. Those are strongly related with the outcomes but may not associate with carbohydrate eating groups may reduce the standard errors and so improve the precisions are also considered in the linear regression models. 
	\item Confounding and/or interaction effect from each of the potential factors were checked one by one. Interaction effects were tested using the adjusted Wald test testing whether the regression coefficients of the interaction terms are simultaneously equal to zero.  
	\item A preliminary model that includes all of the variables suggested to be confounders in the previous step was established. 
	\item The remaining variables were added to the preliminary model one by one to see if any of them may be a confounder in the condition of the presence of the other covariates. 
	\item For logistic regression models (hypertension) under the survey data, goodness-of-fit was assessed using the adapted \code{svylogitgof} command in Stata \parencite{archer2006goodness}. Other diagnostics for regular logistic regression models, such as estimating the pseudo-R\textsuperscript{2},  AIC or BIC, checking the standardized Pearson residuals, or covariate pattern residuals are currently not available for weighted survey data. 
	\item For linear regression models (WC, BMI), the assumption of independent observations is violated as soon as we weighted the sample. General checking such as QQ plots of the residuals (normality), plotting the residuals against fitted values (constant variance) are not available either. Outliers, leverage, and Cook's distance cannot be checked either, however, participants with extreme weightings (if exist) were checked by removing them and refit the models as a sensitivity analysis.
	\item Since under survey designed data, the sampling-weighted least squares are not maximum likelihood, it would not be possible to compare models using likelihood ratio test. Instead, adjusted Wald tests with $p < 0.05$ were used as criteria for variable inclusion in the final model. Another Stata command \code{linktest} was also used to  decide whether quadratic and cubic terms of continuous variables were necessary in improving fitting of the model \parencite{pregibon1980goodness}. 
\end{enumerate}

Data manipulation and preparation \textbf{(Appendix \ref{AppendixA})} were done in R version 3.5.1 \parencite{R3.5.1}. All statistical analyses, except for MLCA models, were performed with \code{svyset} command as implemented in Stata software version 15.1 \parencite{stata15}. The process of model fitting, covariates selection, interaction effect testings for the association between carbohydrate eating patterns and hypertension is shown as an example in \textbf{Appendix \ref{AppendixD}}. All \textit{p} values were two-sided.

>>>>

In the diary, participants were asked to record portion sizes in household measures (e.g. one tablespoon of beans, one Kit Kat finger-size), or for packaged foods to note the weight indicated on the packet. For homemade dishes, participants were asked to record on a separate page in the diary the individual ingredients and quantities for the whole dish along with a brief description of the cooking method and how much of dish they had consumed. For each eating occasion, in addition to the details of what and how much was eaten, participants were also asked to record: when was it, where they were, and who they were eating with. An example, used as guidance for participants, of a food diary for one day is shown in \textbf{Appendix \ref{AppendixE}}.\vspace{-0.3cm}

@@

\begin{itemize}
	\item Total energy intake = (protein $\times$ 17) + (fat $\times$ 37) + (carbohydrate $\times$ 16) + (alcohol $\times$ 29)  kJ;
	\item Carbohydrate intake = total sugars + starch; 
	\item Where all quantities above were measured as mass in grams.
\end{itemize}

Time across a typical survey day was divided into 7 time slots in the dietary diary of NDNS RP: 6 am to 9 am, 9 am to 12 noon, 12 noon to 2 pm, 2 pm to 5 pm, 5 pm to 8 pm, 8 pm to 10 pm, and 10 pm to 6 am next morning. To produce a list of discrete responses for our variable of interest, the energy consumed within each time slot over the four days of survey for each participant were calculated. The percentages of total energy intake contributed by carbohydrate within each time slot were then calculated. Since we planned to apply latent class analysis (LCA) in the current study, in which the observed indicators for latent classes must be categorical, the responses were then dichotomised according to these percentages of the energy intake at a cut-off value of 50\%, i.e. if within a time slot where there any energy intake occurred, carbohydrate consumption was categorised by whether its energy contribution was lower or higher/equal to 50\% of total energy intake within that time slot. Consequently, for each day of the recording, there were 7 data points generated by the diary. Each data point included one of the following responses:

@@

\begin{itemize}
	\item Randomly select PSUs from the Postcode Address File; 
	\item Randomly select 27 or 28 addresses in that postcode area; 
	\item Randomly select one household at that address for interviews; 
	\item Selected addresses where children resided were randomly allocated to one of two groups to determine whether an adult (aged 19 years or older) and a child (aged 1.5 to 18 years) or a child only, were selected for interviews.
\end{itemize}
\vspace{-0.6cm}
%-----------------------------------
%	SUBSECTION 2
%-----------------------------------
\subsection{Response rates}\vspace{-0.3cm}

@@

The strata used to calibrate proportions in the sample include: age-group (1.5-3, 4-6, 7-10, 11-15, 16-18, 19-24, 25-29, 30-39, 40-49, 50-59, 60-64, 65-69, and over 70 years); sex (men or women); and regions (Northern Ireland, Scotland, Wales, and the nine regions of England).

Two steps of the weighting system were designed in the NDNS RP to assure that the combined sample would be representative of the UK population:

\begin{enumerate}
	\item An overall selection weight, which is the product of the address, dwelling unit, catering (household) unit, and individual selection weights, was generated to correct for the unequal selection probabilities. These weights are the inverses of the selection probabilities at each level of the random sampling process, and they can be used to compensate for differences in the chance of selection of an individual.
	\item An iterative procedure was used to adjust the selection weights until the distribution of the weighted sample matched that of the population for age-group, sex, and geographical region. Population distributions were taken from the mid-year population estimates \parencite{OfficeofNationalStat}. 
\end{enumerate}

Another two sets of weights were generated to correct for differential non-response (either due to refusal or inability) to 1) nurse visit and 2) giving a blood sample. Response rates to the nurse visit among those completed a dietary diary was approximately 75\%, to the blood sample in adults were 51\%, 57\%, and 50\% for years 1 to 4, 5 to 6, and 7 to 8, respectively. In creating the nurse/blood sample weight, a logistic regression model was used by the NDNS RP study team to model the relationship between response to nurse visit/giving a blood sample and a set of predictor variables (socio-demographic, participant and catering/household unit characteristics). The model generated a predicted probability for each participant, which is the probability would agree to a nurse visit/provide a blood sample, given the characteristics of the individual and the household unit. Participants with characteristics associated with non-response were under-represented in the sample and therefore receive a low predicted probability. The inverses of these predicted probabilities were used as a set of non-response weights so that participants with a low predicted probability got a larger weight, increasing their representation in the sample. Then the nurse/blood sample weights were re-scaled so that the sum of the weights equalled the number of participants who had a nurse visit or who provided a blood sample. The final nurse/blood weights should, therefore, make the sample participants representative of all eligible persons in the population.

@@

Computer-assisted personal interviews were conducted for the selected individuals by trained interviewers to collect background information on smoking habits (current, ex-smokers, and never), ethnicity (white, non-white), education level (lower than degree/degree or above level), living with a partner or not, and other socio-demographic variables. Participants also had their height, weight, blood pressure, and waist circumferences (WC) measured by the nurses.

Specifically, blood pressure was measured in a sitting position using an automated, validated machine, the Omron HEM907, after a five-minute rest. The mean of the second and third readings, taken at one-minute intervals, were used in the current report. Hypertension was defined as with systolic blood pressure of 140 mmHg or above, and/or diastolic blood pressure of 90 mmHg or above, and/or taking any medication specifically to reduce blood pressure.

@@

Latent class analysis is a statistical technique that identifies categorical latent (unobserved) class variables on the basis of observed categorical variables \parencite{collins2010latent}. It belongs to the family of latent variable models and is directly analogous to the factor analysis model. The major difference is that the latent variable in LCA is categorical, not continuous as in factor analysis. The basic assumptions in LCA are independent observations and local independence, the latter as shown in the fundamental expression of a typical LCA model: \vspace{-0.8cm}

@@

\textbf{Equation \ref{LCA}} indicates that responses for an observation to the measuring variables are independent of one another given its membership in latent class $t$. However, in the NDNS RP data set, the assumption of independent observations is violated. Each individual completed their dietary diary for three/four consecutive days, their diary recordings were later converted into three/four vectors of categorical responses reflecting the type of carbohydrate consumption at each time slot of the day. The observed sequences (observed days) are nested within the participants and therefore are not independent. This nested data structure requires multilevel techniques. 
\vspace{-0.3cm}

@@

Multilevel latent class analysis accounts for the nested structure of the data by allowing latent class intercepts to vary across level 2 units, thereby examining if and how level 2 units influence the level 1 latent classes. These random intercepts allow the probability of membership in a particular level 1 (observation days) latent class to vary across level 2 units (e.g., here in the current context are the individuals). Essentially this allows the probability that an observation day will belong to a
particular day-level latent class to vary across individual-levels.\vspace{-0.5cm}

@@

Thus, for a binary outcome $C_{ij} = 0, 1$ (low $=0$ or high $=1$ carbohydrate eating days), where $i$ denotes the observation days $(i = 1, 2, 3, 4)$, and $j$ denotes the individual $(j = 1, 2, \cdots, 6155)$, the 2-level random intercept logistic regression model can be expressed as:\vspace{-0.4cm}

@@

The same framework can be used to consider random effects in an LCA model, but instead of saying that $C_{ij}$ is either low or high carbohydrate eating days as if we already know the set of latent classes, it is now replaced by a latent variable $G_{ij}$ which indicates the typologies of carbohydrate eating patterns. Then we can use the day-level data to assess the log-odds of belonging to the $k$th type of carbohydrate eating pattern on a specific day of the survey and allow the log-odds to vary across individuals. Therefore, for some persons, the log-odds of having a $k$th type of carbohydrate eating pattern during the survey can be high, but for the other persons, the log-odds of following the $k$th type of carbohydrate eating pattern can be low.

@@

\begin{equation}
P(C_{ij} = t | CB_j = m)  = \frac{\exp(\gamma_{tm})}{\sum_{r=1}^{T}\exp(\gamma_{rm})}
\end{equation}

@@

\begin{itemize}
	\item First, we ignored the multilevel structure of the data and estimated a series of traditional LC models to determine the number of classes at the observational day-level;
	\item Next, a series of MLCA models were fitted to account for the multilevel structure of the data. In these models, the number of day-level classes was based on the best fitting LCA model from the first step, and the LCA model at the individual-level was estimated to identify the number of individual-level latent classes;
	\item Last, once the number of individual-level latent classes was defined based on the previous stage, the number of day-level classes was modified (one class lower and one class higher than in the second step) to investigate the effect of changing level 1 classes and to confirm the best fitting model.
\end{itemize}

The number of classes in level 1 was determined by 1) the evaluation of model fit indices, including the Bayesian information criterion (BIC) and entropy, which is a statistic that summarizes latent class probabilities where values near 1 indicate better latent class separation; 2) the Lo-Mendell-Rubin Likelihood Ratio Test (LMR-LRT) \parencite{lo2001testing, nylund2007deciding} which compares $q$ vs. $q-1$ classes models, where $q$ is the number of latent classes; most importantly, 3) pattern interpretability. In the steps of performing multilevel LCA, where LMR-LRT is not available, the same rules of model fit indices and pattern interpretability were used to determine the optimal combination of latent classes in observation day-level and individual-level. MLCA models were fitted in Mplus 7.4 \parencite{muthen2005mplus}, the Mplus syntax and outputs are shown in \textbf{Appendix \ref{AppendixB}}. \vspace{-0.3cm}

@@

Day-level latent classes identified by the first step of MLCA were tabulated according to the day of week and also whether the diary was recorded during weekends or not. A contingency table giving the frequency of responses across the 7-time slots of the survey days was produced. Descriptive statistics for the dietary day-level recordings according to the latent class memberships were presented. Pearson $\chi^2$ test was used to assess evidence fora difference in the distribution of categorical variables. One-way Analysis of Variance (ANOVA) was used to compare the means across the multiple groups for continuous variables.

@@

\begin{enumerate}
	\item The crude association between the carbohydrate eating groups and the outcomes was first examined. 
	\item Potential confounders of the association between carbohydrate eating groups (exposure) and the outcomes were selected depending on the descriptive statistical analyses conducted above, i.e. those associated with both the exposure and the outcome and also not on the causal pathway were selected as potential confounders. Covariates that are strongly related with the outcomes but not associate with carbohydrate eating groups may reduce the standard errors and improve the precisions, so they were also considered in the linear regression models. 
	\item Confounding and/or interaction effects from each of the potential factors were checked one by one. Interaction effects were tested using the adjusted Wald test for whether the regression coefficients of the interaction terms are simultaneously equal to zero.  
	\item A preliminary model that includes all of the variables suggested to be confounders in the previous step was established. 
	\item The remaining variables were added to the preliminary model one by one to see if any of them may be a confounder in the condition of the presence of the other covariates. 
	\item For logistic regression models (hypertension) under the survey data, goodness-of-fit was assessed using the adapted \code{svylogitgof} command in Stata \parencite{archer2006goodness}. Other diagnostics for regular logistic regression models, such as estimating the pseudo-R\textsuperscript{2},  AIC or BIC, checking the standardized Pearson residuals, or covariate pattern residuals are currently not available for weighted survey data. 
	\item For linear regression models (WC, BMI), the assumption of independent observations is violated as soon as we weight the sample. General checking such as QQ plots of the residuals (normality), plotting the residuals against fitted values (constant variance) are not available either. Outliers, leverage, and Cook's distance cannot be checked either, however, participants with extreme weightings (if any) were checked by removing them and refitting the models as a sensitivity analysis.
	\item Since under survey designed data, the sampling-weighted least squares are not maximum likelihood, it would not be possible to compare models using likelihood ratio test. Instead, adjusted Wald tests with $p < 0.05$ were used as criteria for variable inclusion in the final model. Another Stata command \code{linktest} was also used to  decide whether quadratic and cubic terms of continuous variables were necessary in improving fitting of the model \parencite{pregibon1980goodness}. 
\end{enumerate}

Data manipulation and preparation \textbf{(Appendix \ref{AppendixA})} were done in R version 3.5.1 \parencite{R3.5.1}. All statistical analyses, except for MLCA models, were performed with \code{svyset} command as implemented in Stata software version 15.1 \parencite{stata15}. The process of model fitting, covariate selection, and interaction effect testing for the association between carbohydrate eating patterns and hypertension is shown as an example in \textbf{Appendix \ref{AppendixD}}. All \textit{p} values were two-sided.
